---
title: "Automatic Model Simplification in sdmTMB: Collapsing random fields and simplifying autoregressive structures"
author: "Andrew Allyn"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE,
  fig.width = 7,
  fig.height = 5
)
devtools::load_all()
```

# Overview

## Motivation

Complex spatial and spatiotemporal models can sometimes estimate parameters that suggest the model is overparameterized. This work builds on [Eric Ward's `collapse_random_fields` contribution](https://github.com/sdmTMB/sdmTMB/pull/477), which addressed issues with spatial and spatiotemporal random field variances collapsing to zero. 

In addition to these variance-related issues, we commonly encounter problems with autoregressive (AR1) temporal correlation parameters. When fitting AR1 processes, the estimated correlation parameter (rho or rho_time for the spatiotemporal AR1 process) can sometimes indicate that:

1. **High correlation (rho or rho_time ≈ 1):** The process is essentially a random walk, suggesting `"rw"` is more appropriate than `"ar1"`
2. **No correlation (rho or rho_time ≈ 0):** The process shows no temporal correlation, suggesting simpler structures like `"iid"` or random effects

Taking inspiration from Eric's function, I developed a unified `check_and_collapse()` function that automatically detects and addresses these issues.

## What the Function Detects

The unified collapse function checks for:

### Variance Issues (replicating functioality from `check_and_collapse_random_fields`)
- **Spatial variance → 0:** `sigma_O` very small → collapse `spatial = "on"` to `"off"`
- **Spatiotemporal variance → 0:** `sigma_E` very small → collapse `spatiotemporal = "iid/ar1/rw"` to `"off"`

### Correlation Issues (AR1 processes)
- **Spatiotemporal AR1 → RW:** `rho` approaching 1 → collapse `spatiotemporal = "ar1"` to `"rw"`
- **Spatiotemporal AR1 → IID:** `rho` approaching 0 → collapse `spatiotemporal = "ar1"` to `"iid"`
- **Time-varying AR1 → RW:** `rho_time` approaching 1 → collapse `time_varying_type = "ar1"` to `"rw"`
- **Time-varying AR1 → IID:** `rho_time` approaching 0 → **warning only** (user must manually change formula)

### Detection Criteria

For each parameter, we use **two checks with OR logic**:

1. **Point estimate:** Is the estimate within the threshold of the boundary?
2. **Confidence interval:** Does the 95% CI overlap the boundary?

If **either** condition is met, the function triggers collapse/warning.

**Example for low correlation:**
- Threshold = 0.01
- Check 1: Is |rho| < 0.01?
- Check 2: Does 95% CI include 0?
- If YES to either → Collapse to IID (or warn)

This OR logic is more robust to noisy parameter estimates and helps identify when the model is uncertain about the parameter value.

## Implementation

### The `check_and_collapse()` Function

This internal function:

1. Examines fitted parameter estimates and their uncertainty
2. Determines if parameters are at boundary values
3. Returns updated model structure specifications
4. Stores warnings for display after fitting completes

### Integration in `fit.R`

The collapse checking is integrated into the main `sdmTMB()` fitting function:

1. Initial model is fit
2. If collapse checking is enabled, `check_and_collapse()` examines the fit
3. If issues are detected:
   - For auto-fixable issues (spatial variance, spatiotemporal variance/correlation): Model is automatically refit
   - For manual-fix issues (time-varying AR1 → IID): Warning is displayed after fitting
4. Final model is returned

### Control Parameters

```r
sdmTMBcontrol(
  collapse = TRUE,                        # What to check
  collapse_variance_threshold = 0.01,     # Sigma threshold
  collapse_correlation_threshold = 0.01   # Rho threshold
)
```

**`collapse` options:**
- `TRUE` (default): Check both variance and correlation
- `FALSE`: Disable all checking
- `"variance"`: Check variance only (sigma_O, sigma_E)
- `"correlation"`: Check correlation only (rho, rho_time)

---

# Example Tests

We'll demonstrate the collapse functionality using simulated data where we know the true structure.

## Setup

```{r setup-data}
set.seed(123)

# Create spatial grid
predictor_dat <- data.frame(
  X = runif(1000), 
  Y = runif(1000),
  a1 = rnorm(1000), 
  year = sample(1:5, size = 1000, replace = TRUE)
)

mesh <- make_mesh(predictor_dat, xy_cols = c("X", "Y"), cutoff = 0.1)
```

---

## Test 1: Variance Collapse (Spatial and Spatiotemporal)

**Scenario:** Simulate data with NO spatial or spatiotemporal variation (sigma_O = 0, sigma_E = 0), then fit a model that includes these fields.

### Problem: Without Collapse Detection

```{r test1-problem}
# Simulate with zero variance
sim_dat <- sdmTMB_simulate(
  formula = ~ 1 + a1,
  data = predictor_dat,
  time = "year",
  mesh = mesh,
  family = tweedie(),
  range = 2,
  sigma_E = 0.005,      # Near zero spatiotemporal variation
  sigma_O = 0.005,      # Near zero spatial variation
  phi = 0.2,
  seed = 42,
  B = c(0.2, -0.4)
)

# Fit model with collapse disabled
fit1_problem <- sdmTMB(
  observed ~ a1,
  data = sim_dat, 
  mesh = mesh, 
  time = "year",
  spatial = "on",           # Will estimate near-zero
  spatiotemporal = "iid",   # Will estimate near-zero
  control = sdmTMBcontrol(collapse = FALSE)
)

# Check estimated variances
cat("Spatial field:", fit1_problem$spatial, "\n")
cat("Spatiotemporal field:", fit1_problem$spatiotemporal, "\n")
cat("Estimated sigma_O:", fit1_problem$tmb_obj$report()$sigma_O, "\n")
cat("Estimated sigma_E:", fit1_problem$tmb_obj$report()$sigma_E, "\n")
```

**Issue:** The model includes unnecessary random fields that add complexity without improving fit, with both `sigma_O` and `sigma_E` collapsing to 0.

### Solution: With Collapse Detection

```{r test1-solution}
# Fit model with collapse enabled (default)
fit1_solution <- sdmTMB(
  observed ~ a1,
  data = sim_dat, 
  mesh = mesh, 
  time = "year",
  spatial = "on",
  spatiotemporal = "iid",
  control = sdmTMBcontrol(collapse = TRUE)
)

# Check updated structure
cat("Spatial field:", fit1_solution$spatial, "\n")
cat("Spatiotemporal field:", fit1_solution$spatiotemporal, "\n")
```

**Result:** Both fields automatically collapsed to `"off"`, simplifying the model.

---

## Test 2: High Correlation (AR1 → RW)

**Scenario:** Simulate time-varying intercept with very high temporal correlation (rho ≈ 1), then fit with AR1 structure.

### Problem: Without Collapse Detection

```{r test2-setup}
# Likely need some more time steps for this one
set.seed(123)
predictor_dat_tv <- data.frame(
  X = runif(1000), 
  Y = runif(1000),
  a1 = rnorm(1000), 
  year = rep(1:20, length.out = 1000)  # 15 time points for better AR1 estimation
)

mesh_tv <- make_mesh(predictor_dat_tv, xy_cols = c("X", "Y"), cutoff = 0.1)

# Simulate with rho_time = 0.95 (very high correlation)
sim_dat_high_rho <- sdmTMB_simulate(
  formula = ~ 1 + a1,
  data = predictor_dat_tv,
  time = "year",
  mesh = mesh_tv,
  family = gaussian(),
  range = 0.5,
  time_varying = ~ 1,
  time_varying_type = "ar1",
  rho_time = 0.99,        # High correlation
  sigma_V = 0.8,          # Make it larger so it's noticeable!
  sigma_E = 0,
  sigma_O = 0,
  phi = 0.1,
  seed = 99,
  B = c(1, -0.1)
)
```

```{r test2-problem}
# Fit with collapse disabled
fit2_problem <- sdmTMB(
  observed ~ 1 + a1,
  data = sim_dat_high_rho,
  mesh = mesh,
  time = "year",
  spatial = "off",
  spatiotemporal = "off",
  time_varying = ~ 1,
  time_varying_type = "ar1",
  control = sdmTMBcontrol(collapse = FALSE)
)

# Check estimated rho
pl <- as.list(fit2_problem$sd_report, "Estimate")
pls <- as.list(fit2_problem$sd_report, "Std. Error")
rho_unscaled <- pl$rho_time_unscaled[1,1]
se_unscaled <- pls$rho_time_unscaled[1,1]

# Transform to bounded scale
bound <- function(x) 2 * plogis(x) - 1
rho_est <- bound(rho_unscaled)
rho_lower <- bound(rho_unscaled - 1.96 * se_unscaled)
rho_upper <- bound(rho_unscaled + 1.96 * se_unscaled)

cat("Estimated rho:", round(rho_est, 3), "\n")
cat("95% CI: [", round(rho_lower, 3), ",", round(rho_upper, 3), "]\n")
cat("Time-varying type:", as.character(fit2_problem$call['time_varying_type']), "\n")
```

**Issue:** AR1 is unnecessarily complex when rho ≈ 1; a random walk would be simpler and equally appropriate.

### Solution: With Collapse Detection

```{r test2-solution}
# Fit with collapse enabled
fit2_solution <- sdmTMB(
  observed ~ 1 + a1,
  data = sim_dat_high_rho,
  mesh = mesh,
  time = "year",
  spatial = "off",
  spatiotemporal = "off",
  time_varying = ~ 1,
  time_varying_type = "ar1",
  control = sdmTMBcontrol(collapse = TRUE,
    collapse_correlation_threshold = 0.05)
)

cat("Time-varying type:", as.character(fit2_solution$call['time_varying_type']), "\n")
```

**Result:** AR1 automatically switched to RW (random walk).

---

## Test 3: Low Correlation (AR1 → IID/Warning)

**Scenario:** Simulate time-varying effects with no temporal correlation (rho = 0), then fit with AR1 structure.

### Problem: Without Collapse Detection

```{r test3-setup}
# Simulate with rho = 0 (no temporal correlation)
sim_dat_low_rho <- sdmTMB_simulate(
  formula = ~ 1 + a1,
  data = predictor_dat_tv,
  time = "year",
  mesh = mesh_tv,
  family = gaussian(),
  range = 0.5,
  time_varying = ~ 1,
  time_varying_type = "ar1",
  rho_time = 0.01,              # No correlation
  sigma_V = 0.8,
  sigma_E = 0,
  sigma_O = 0,
  phi = 0.1,
  seed = 99,
  B = c(1, -0.1)
)
```

```{r test3-problem}
# Fit with collapse disabled
fit3_problem <- sdmTMB(
  observed ~ 1 + a1,
  data = sim_dat_low_rho,
  mesh = mesh_tv,
  time = "year",
  spatial = "off",
  spatiotemporal = "off",
  time_varying = ~ 1,
  time_varying_type = "ar1",
  control = sdmTMBcontrol(collapse = FALSE)
)

# Check estimated rho
pl <- as.list(fit3_problem$sd_report, "Estimate")
pls <- as.list(fit3_problem$sd_report, "Std. Error")
rho_unscaled <- pl$rho_time_unscaled[1,1]
se_unscaled <- pls$rho_time_unscaled[1,1]

rho_est <- bound(rho_unscaled)
rho_lower <- bound(rho_unscaled - 1.96 * se_unscaled)
rho_upper <- bound(rho_unscaled + 1.96 * se_unscaled)

cat("Estimated rho:", round(rho_est, 3), "\n")
cat("95% CI: [", round(rho_lower, 3), ",", round(rho_upper, 3), "]\n")
cat("CI overlaps 0:", rho_lower < 0 & rho_upper > 0, "\n")
```

**Issue:** AR1 is overparameterized when there's no temporal correlation. The model is also uncertain about rho (wide CI crossing 0).

### Solution: With Collapse Detection

```{r test3-solution}
# Fit with collapse enabled
fit3_solution <- sdmTMB(
  observed ~ 1 + a1,
  data = sim_dat_low_rho,
  mesh = mesh_tv,
  time = "year",
  spatial = "off",
  spatiotemporal = "off",
  time_varying = ~ 1,
  time_varying_type = "ar1",
  control = sdmTMBcontrol(collapse = TRUE),
  silent = FALSE
)

cat("Time-varying type:", as.character(fit3_solution$call['time_varying_type']), "\n")
```

**Result:** A warning is displayed suggesting the user remove `time_varying` or use `(1 | year)` instead. The model continues fitting with AR1 as specified, but the user is informed about the issue.

**Note:** For `time_varying`, the function cannot automatically switch to "iid" because `time_varying_type = "iid"` is not a valid option in sdmTMB. The user must manually change the formula to use random effects: `formula = observed ~ 1 + (1 | year)`.

---

## Test 4: Selective Checking (variance only)

**Scenario:** Check only variance, not correlation.

```{r test4}
# Fit checking variance only
fit4 <- sdmTMB(
  observed ~ 1 + a1,
  data = sim_dat_low_rho,  # Has low rho AND low spatial variance
  mesh = mesh_tv,
  time = "year",
  spatial = "on",           # Low variance - will collapse
  spatiotemporal = "off",
  time_varying = ~ 1,
  time_varying_type = "ar1",  # Low rho - will NOT be checked
  control = sdmTMBcontrol(collapse = "variance"),
  silent = FALSE
)

cat("Spatial:", fit4$spatial, "\n")
cat("Time-varying type:", as.character(fit4$call['time_varying_type']), "\n")
```

**Result:** Spatial field collapsed (variance check), but AR1 remains unchanged (correlation not checked) AND we don't see the warning because the correlation check was turned off.

---

## Test 5: Selective Checking (correlation only)

**Scenario:** Check only correlation, not variance.

```{r test5}
# Fit checking variance only
fit5 <- sdmTMB(
  observed ~ 1 + a1,
  data = sim_dat_low_rho,  # Has low rho AND low spatial variance
  mesh = mesh_tv,
  time = "year",
  spatial = "on",           # Low variance - but not checked
  spatiotemporal = "off",
  time_varying = ~ 1,
  time_varying_type = "ar1",  # Low and poorly estimated - will be checked and give warning
  control = sdmTMBcontrol(collapse = "correlation"),
  silent = FALSE
)

cat("Spatial:", fit5$spatial, "\n")
cat("Time-varying type:", as.character(fit5$call['time_varying_type']), "\n")
```

**Result:** Spatial field remains unchanged, and a warning is displayed suggesting the user remove `time_varying` or use `(1 | year)` instead. The model continues fitting with AR1 as specified
---

## Test 6: Exhaustive Checking (variance and correlation)

**Scenario:** Check variance and corrrelation

```{r test6}
# Fit checking variance and correlation
fit6 <- sdmTMB(
  observed ~ 1 + a1,
  data = sim_dat_low_rho,  # Has low rho AND low spatial variance
  mesh = mesh_tv,
  time = "year",
  spatial = "on",           # Low variance - will collapse
  spatiotemporal = "off",
  time_varying = ~ 1,
  time_varying_type = "ar1",  # Low and poorly estimated rho - will be checked and give warning
  control = sdmTMBcontrol(collapse = TRUE),
  silent = FALSE
)

cat("Spatial:", fit6$spatial, "\n")
cat("Time-varying type:", as.character(fit6$call['time_varying_type']), "\n")
```

**Result:** Spatial field collapsed (variance check), and a warning is displayed suggesting the user remove `time_varying` or use `(1 | year)` instead. The model continues fitting with AR1 as specified
---

## Test 7: Custom thresholds

**Scenario:** Adjust variance threshold

```{r test7}
sim_dat_small_var <- sdmTMB_simulate(
  formula = ~ 1 + a1,
  data = predictor_dat,
  time = "year",
  mesh = mesh,
  family = tweedie(),
  range = 2,
  sigma_E = 0,    # Small but not tiny
  sigma_O = 0.05,
  phi = 0.2,
  seed = 45,
  B = c(1, -0.1)
)

# Need to fit the model to get some idea of the threshold for demonstration purposes
fit7_demo <- sdmTMB(
  observed ~ 1 + a1,
  data = sim_dat_small_var,  
  mesh = mesh_tv,
  time = "year",
  spatial = "on",           
  spatiotemporal = "off", 
  control = sdmTMBcontrol(collapse = FALSE),
  silent = FALSE
)

cat("Estimated sigma_O:", fit7_demo$tmb_obj$report()$sigma_O, "\n")

# Should by default collapse, unless threshold is increased
thresh_use<- as.numeric(fit7_demo$tmb_obj$report()$sigma_O) + 0.1

fit7 <- sdmTMB(
  observed ~ 1 + a1,
  data = sim_dat_small_var,  
  mesh = mesh_tv,
  time = "year",
  spatial = "on",           
  spatiotemporal = "off",
  control = sdmTMBcontrol(collapse = TRUE,
    collapse_variance_threshold = thresh_use),
  silent = FALSE
)

cat("Spatial:", fit7$spatial, "\n")
```

**Result:** Spatial field collapsed (variance check) with custom threshold, when it wouldn't under default (collapse_variance_threshold = 0.01).
---

# Summary

## Key Features

1. **Unified detection:** Single function checks both variance and correlation issues
2. **Robust criteria:** Uses both point estimates and confidence intervals
3. **Flexible control:** User can enable/disable specific checks
4. **Automatic refitting:** Model simplification happens automatically when possible
5. **Clear warnings:** Issues that require manual intervention are clearly communicated

## Best Practices

**When to use collapse checking:**
- Exploratory data analysis
- Model selection
- When you're unsure about appropriate random effect structure
- Production models where automatic simplification is desired

**When to disable collapse checking:**
- When you want exact control over model structure
- Sensitivity analyses
- When testing specific hypotheses about variance/correlation parameters

## Control Parameter Guide

```r
# Default: Check everything
sdmTMBcontrol(collapse = TRUE)

# Disable all checks
sdmTMBcontrol(collapse = FALSE)

# Only check variance
sdmTMBcontrol(collapse = "variance")

# Only check correlation
sdmTMBcontrol(collapse = "correlation")

# Custom thresholds
sdmTMBcontrol(
  collapse = TRUE,
  collapse_variance_threshold = 0.05,     # More aggressive
  collapse_correlation_threshold = 0.02   # More conservative
)
```

## What Gets Checked

| Component | Parameter | Collapse to | Auto-refit? |
|-----------|-----------|-------------|-------------|
| Spatial field | sigma_O | "off" | ✅ Yes |
| Spatiotemporal field | sigma_E | "off" | ✅ Yes |
| Spatiotemporal AR1 (high rho) | rho → 1 | "rw" | ✅ Yes |
| Spatiotemporal AR1 (low rho) | rho → 0 | "iid" | ✅ Yes |
| Time-varying AR1 (high rho) | rho_time → 1 | "rw" | ✅ Yes |
| Time-varying AR1 (low rho) | rho_time → 0 | Warning | ❌ No |

---

# Acknowledgments

This work builds on Eric Ward's `collapse_random_fields` contribution. The extension to handle autoregressive correlation parameters was developed to address common model overparameterization issues encountered in practice.

---

# Session Info

```{r session-info}
sessionInfo()
```

